---
title: ''
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidyr)
library(readxl)
library(janitor)
library(lubridate)
library(gganimate)
library(transformr)
library(ggplot2)
library(stringr)
library(gganimate)
library(gifski)
library(skimr)
library(GGally)
library(tidygeocoder)
library(caret)
library(ROCR)
library(pROC)

install.packages('ROCR')

# Load the data 
tcc <- read.csv("ttc-bus-delay-data-2022.csv")

# Clean the data 
tcc$Date <- as.Date(tcc$Date, format = "%Y-%m-%d")
tcc$Time <- as.POSIXct(strptime(tcc$Time, format = "%H:%M:%S"), 
                       format = "%Y-%m-%d %H:%M:%S", tz = "UTC")
tcc <- clean_names(tcc)
tcc <- tcc %>%
  filter(!str_detect(route, "^\\s*$"))
```

#### Toronto's Public Transportation System
## The Toronto Transit Commission (TTC)

The Toronto Transit Commission (TTC) is Toronto's public transportation system. It is a vast network that covers Toronto and the surrounding municipalities, which are collectively known as the Greater Toronto Area (GTA). The TTC operates subways, buses, and streetcars throughout the region, making it possible to travel to almost any destination. The bus system is particularly extensive, with routes that reach almost every corner of the city. The TTC is one of the fastest and most convenient ways to get around Toronto and the GTA, and I highly recommend it to anyone visiting or living in the area.

```{r cars, warning=FALSE}
skim(tcc)
```



## Is a particular bus route that is more late than others?

```{r pressure, echo=FALSE}
# Is a particular route more late than others?
route_mean_delays <- tcc %>%
  group_by(route) %>%
  summarise(MeanDelay = mean(min_delay, na.rm = TRUE)) %>%
  arrange(desc(MeanDelay)) %>%
  head(10)

# Create a bar plot to display the routes with the 10 highest mean delays
ggplot(route_mean_delays, aes(x = reorder(route,MeanDelay,decreasing=FALSE), y = MeanDelay)) +
  geom_bar(stat = "identity", fill = "red", color = "blue") +
  labs(
    title = "Top 10 Routes with Highest Mean Delay",
    x = "Route",
    y = "Mean Delay (Minutes)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
```



## Is a particular bus route that is more on time than others?

```{r pressure2, echo=FALSE}
# Calculate the mean delay for each route
all_route_delay <- tcc %>%
  group_by(route) %>%
  summarise(MeanDelay = mean(min_delay))

# Sort routes by mean delay in descending order
all_route_delay_summary <- all_route_delay %>%
  arrange(desc(MeanDelay))


# what routs are on time the most often 
route_mean_on_time <- tcc %>%
  group_by(route) %>%
  summarise(MeanDelay = mean(min_delay, na.rm = FALSE)) %>%
  arrange(MeanDelay) %>%
  head(10)

# Create a bar plot to display the routes with the 10 lowest mean delays
#
ggplot(route_mean_on_time, aes(x = reorder(route,MeanDelay,decreasing=TRUE),route, y = MeanDelay)) +
  geom_bar(stat = "identity", fill = "lightgreen", color = "darkgreen") +
  labs(
    title = "Top 10 Routes with Lowest Mean Delay (Most On-Time Routes)",
    x = "Route",
    y = "Mean Delay (Minutes)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels
```


# Citations:

Data source: https://www.kaggle.com/datasets/reihanenamdari/toronto-bus-delay-2022

[1] https://en.wikipedia.org/wiki/Public_transportation_in_Toronto
[2] https://arrivein.com/daily-life-in-canada/getting-around-how-to-use-public-transportation-in-toronto/
[3] https://www.prepareforcanada.com/choosing-a-city/toronto/public-transportation-system-in-torontoride-the-ttc/
[4] https://apollocover.com/magazine/toronto-public-transportation
[5] https://en.wikipedia.org/wiki/Toronto_Transit_Commission_bus_system
[6] https://travel.usnews.com/Toronto/Getting_Around/
